Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-12 dev/paper/paper.tex
almost done, findings still need to be fleshed out though

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

Recent advances in Artificial Intelligence (AI) have propelled its
adoption in scientific domains outside of Computer Science including
Healthcare, Bioinformatics, Genetics and the Social Sciences. While this
has in many cases brought benefits in terms of efficiency,
state-of-the-art models like Deep Neural Networks (DNN) have also given
rise a new type of principal-agent problem in the context of data-driven
decision-making. It involves a group of \textbf{principals} - i.e.~human
stakeholders - that fail to understand the behaviour of their
\textbf{agent} - i.e.~the model used for automated decision-making
\protect\hyperlink{ref-borch2022machine}{{[}1{]}}.

Models or algorithms that fall into this category are typically referred
to \textbf{black-box} models. Despite their shortcomings, black-box
models have grown in popularity in recent years and have at times
created undesirable societal outcomes
\protect\hyperlink{ref-o2016weapons}{{[}2{]}}. The scientific community
has tackled this issue from two different angles: while some have
appealed for a strict focus on inherently interpretable models
\protect\hyperlink{ref-rudin2019stop}{{[}3{]}}, others have investigated
different ways to explain the behaviour of black-box models. These two
sub-domains can be broadly referred to as \textbf{interpretable AI} and
\textbf{explainable AI} (XAI), respectively.

@@

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}}. In this work, we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

@@

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself. The term \textbf{macrodynamics} is borrowed from the
economics literature and used to describe processes involving whole
groups or societies.

@@

\caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
simple Bayesian model trained for binary classification (a); the
implementation of AR for a random subset of individuals leads to a
domain shift (b); as the classifier is retrained we observe a model
shift (c); as this process is repeated, the decision boundary moves away
from the target class (d).}

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on Algorithmic Recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of Algorithmic Recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} posits a generalized
framework for gradient-based Algorithmic Recourse and introduces the
notion of hidden external costs. Section~\ref{sec-method-2} sets out our
experimental framework for modeling endogenous macrodynamics in AR.
Section~\ref{sec-empirical} presents our experimental results for
synthetic and real-world datasets. We also present evidence for the
effectiveness of various proposed mitigation strategies in that section.
Our findings are then discussed in the broader context of the literature
in Section~\ref{sec-discussion}. We also point to some of the
limitations of our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
cost\footnote{Equivalently, others have referred to this quantity as
  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
by individual \(x\) when moving to a counterfactual state \(x\prime\)
such that the predicted outcome \(M(x\prime)\) corresponds to some
target outcome \(y^*\):

@@

In the baseline work
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, the cost
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that perturbations of
\(x\) are costly to the individual. For models that are differentiable
and produce smooth predictions, Equation~\ref{eq-solution} can be solved
through gradient descent. This summarizes the approach followed in
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} which we
shall refer to simply as \textbf{Wachter} - the name of the first author
- in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been
described in the literature since 2017. An October 2020 survey by Karimi
et al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}11{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}13{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}14{]}}.

Efforts so far have largely been directed at improving the quality of
Counterfactual Explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
interested in generating one or multiple meaningful Counterfactual
Explanations for some individual characterized by \(x\). The ability of
Counterfactual Explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}12{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} propose ROAR - a
framework for Algorithmic Recourse that evidently improves robustness to
such exogenous shifts.

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{{[}16{]}}. The authors
propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}17{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}18{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}19{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}20{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}21{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}22{]}}.

In the context of Algorithmic Recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}15{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search, there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}23{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework - CARLA - that can be used to apply and benchmark different
methodologies; finally, \texttt{CounterfactualExplanations.jl}
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits. In particular, we have built a
framework on top of \texttt{CounterfactualExplanations.jl} that extends
the functionality from static benchmarks to simulation experiments:
\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
  available from \ldots{}}. The core concepts implemented in that
package reflect what is presented in section Section~\ref{sec-method-2}
of this paper.

@@

We have chosen to focus on gradient-based counterfactual search for two
reasons: firstly, they can be seen as direct descendants of our baseline
method - Wachter; secondly, gradient-based search is particularly
well-suited for differentiable black-box models like deep neural
networks, which we focus on in this work. In particular, we include
include the following generators in our simulation experiments below:
\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}10{]}},
\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
and a greedy approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}7{]}}. Our motivation for
including these different generators in our analysis, is that they all
offer slightly different approaches to generate meaningful
counterfactuals for differentiable black-box models. We hypothesize that
generating more \textbf{meaningful} counterfactuals should mitigate the
endogenous dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro}. This intuition stems from the underlying idea
that more meaningful counterfactuals are generated by the same or at
least a very similar data generating process as the training data. All
else equal, counterfactuals that fulfill this basic requirement should
be less prone to trigger domain and model shifts.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In Wachter, the state space is the
feature space: \(f\) is just the identity function and the number of
counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
instead of the feature space directly. The latent embedding is learned
by a separate generative model that is tasked with learning the data
generating process (DGP) of \(X\). In this case \(f\) in
Equation~\ref{eq-general} corresponds to the decoder part of the
generative model, in other words the function that maps back from the
latent embedding to the feature space. Provided the generative model is
well-specified, traversing the latent embedding typically results in
realistic and plausible counterfactuals, because they are implicitly
generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}}.

CLUE distinguishes itself from REVISE and other counterfactual
generators in that it aims to minimize the predictive uncertainty of the
model in question, \(M\). To quantify predictive uncertainty the authors
rely on entropy estimates for probabilistic models. The approach
proposed by \protect\hyperlink{ref-schut2021generating}{{[}7{]}}, which
we shall refer to as \textbf{Greedy}, also works with the subclass of
models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
predictive uncertainty estimates. The authors show that in this setting
the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
and meaningful counterfactuals can be generated in a fast and efficient
manner through a modified Jacobian-based Saliency Map Attack (JSMA).
Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}7{]}} also
show that by maximizing the predicted probability of \(x^\prime\) being
assigned to target class \(y^*\) we also implicitly minimize predictive
entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
REVISE in the Bayesian context and we shall therefore refer to both
approaches collectively as \textbf{Latent Space} generators\footnote{In
  fact, there are a number of other recently proposed approaches to
  counterfactual search that also broadly fall in this same category.
  They largely differ with respect to the chosen generative model: for
  example, the generator proposed by
  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}24{]}} relies
  on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}} show that
diverse outcomes can in principal be achieved simply rerunning
counterfactual search multiple times using stochastic gradient descent
(or by randomly initializing the counterfactual). In
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} diversity is
explicitly proxied via Determinantal Point Processes (DDP): the authors
simply introduce DDP as a component of the cost function
\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
\(s_1, ... , s_K\) that look as different from each other as possible.
The implementation of DiCE in the our library of choice -
\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
constant and therefore does not affect the objective function
Equation~\ref{eq-general}.

@@

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting and focused
on single individuals: given some estimated model the goal is to find
valid counterfactuals for individual instance that fulfill various
desiderata. The ability of such counterfactuals to handle dynamics like
data and model drift remains a largely unexplored research challenge at
this point. There has also been surprisingly little work on the related
question of how the actual implementation of recourse by one individual
may affect other individuals. Through this work we aim to close that gap
by systematizing and extending existing knowledge. We first show that
many of the existing methodologies can be collectively described by a
generalized framework. We then argue that the existing framework fails
to account for a hidden external cost of recourse, that only reveals
itself when studying the endogenous dynamics of recourse at the group
level. Through simulation experiments involving various popular
counterfactual generators and several benchmark datasets, we generate a
total XX million Counterfactual Explanations and study the resulting
domain and model shifts. We find that the induced shifts are substantial
enough to likely impede the applicability Algorithmic Recourse in
practice. Fortunately, we find various potential mitigation strategies
that can be used in combination with existing approaches. Our simulation
framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

Recent advances in Artificial Intelligence (AI) have propelled its
adoption in scientific domains outside of Computer Science including
Healthcare, Bioinformatics, Genetics and the Social Sciences. While this
has in many cases brought benefits in terms of efficiency,
state-of-the-art models like Deep Neural Networks (DNN) have also given
rise a new type of problem in the context of data-driven
decision-making. They are essentially \textbf{black boxes}: so complex,
opaque and underspecified in the data that it is often impossible to
understand how they actually arrive at their decision without auxiliary
tools. Despite this shortcoming, black-box models have grown in
popularity in recent years and have at times created undesirable
societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}. The
scientific community has tackled this issue from two different angles:
while some have appealed for a strict focus on inherently interpretable
models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have
investigated different ways to explain the behaviour of black-box
models. These two sub-domains can be broadly referred to as
\textbf{interpretable AI} and \textbf{explainable AI} (XAI),
respectively.

@@

Existing work in this field has largely worked in a static setting:
various approaches have been proposed to generate counterfactuals for a
given individual that is subject to some pre-trained model. More recent
work has compared different approaches within this static setting
\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}. In this work, we go
one step further and ask ourselves: what happens if recourse is provided
and implemented repeatedly? What types of dynamics are introduced and
how do different counterfactual generators compare in this context?

@@

Figure~\ref{fig-poc} illustrates this idea for a binary problem
involving a probabilistic classifier and the counterfactual generator
proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}:
the implementation of AR for a subset of individuals leads to a domain
shift (b), which in turn triggers a model shift (c). As this game of
implementing AR and updating the classifier is repeated, the decision
boundary moves away from training samples that were originally in the
target class (d). We refer to these types of dynamics as
\textbf{endogenous} because they are induced by the implementation of
recourse itself. The term \textbf{macrodynamics} is borrowed from the
economics literature and used to describe processes involving whole
groups or societies.

@@

\caption{\label{fig-poc}Dynamics in Algorithmic Recourse: we have a
simple linear classifier trained for binary classification (a); the
implementation of AR for a random subset of individuals leads to a
noticable domain shift (b); as the classifier is retrained we observe a
corresponding model shift (c); as this process is repeated, the decision
boundary moves away from the target class (d).}

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on Algorithmic Recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}},
which enables us to study macrodynamics of Algorithmic Recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}},
\protect\hyperlink{ref-schut2021generating}{{[}6{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}9{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

The remainder of the paper is structured as follows:
Section~\ref{sec-related} places our work in the broader context of
related literature. Section~\ref{sec-method} posits a generalized
framework for gradient-based Algorithmic Recourse and introduces the
notion of hidden external costs. Section~\ref{sec-method-2} sets out our
experimental framework for modeling endogenous macrodynamics in AR.
Section~\ref{sec-empirical} presents our experimental setup, including
the data, classifiers and counterfactual generators that we have
employed. Section~\ref{sec-empirical-2} presents the results for
synthetic and real-world datasets along with evidence for the
effectiveness of various proposed mitigation strategies. Our findings
are then discussed in the broader context of the literature in
Section~\ref{sec-discussion}, before pointing to some of the limitations
of our work as well as avenues for future research in
Section~\ref{sec-limit}. Finally, Section~\ref{sec-conclusion}
concludes.

@@

A framework for Counterfactual Explanations was first proposed in 2017
by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has
served as the baseline for most methodologies that have been proposed
since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some
pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs
\(Y \in \mathcal{Y}\). Then we are interested in minimizing the
cost\footnote{Equivalently, others have referred to this quantity as
  \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred
by individual \(x\) when moving to a counterfactual state \(x\prime\)
such that the predicted outcome \(M(x\prime)\) corresponds to some
target outcome \(y^*\):

@@

In the baseline work
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, the cost
function \(h: \mathcal{X} \mapsto \mathbb{R}\) is proxied by some
distance metric based on the simple intuition that perturbations of
\(x\) are costly to the individual. For models that are differentiable
and produce smooth predictions, Equation~\ref{eq-solution} can be solved
through gradient descent. This summarizes the approach followed in
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we
shall refer to simply as \textbf{Wachter} - the name of the first author
- in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been
described in the literature since 2017. An October 2020 survey by Karimi
et al.~layed out 60 algorithms that have been proposed since 2014
\protect\hyperlink{ref-karimi2020survey}{{[}10{]}}. Another survey
published around the same time by Verma et al.~described 29 algorithms
\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. Different
approaches vary primarily in terms of the objective functions they
impose, how they optimize said objective (from brute force through
gradient-based approaches to graph traversal algorithms), and how the
ensure that certain requirements for CE are met. Regarding the latter,
the literature has produced an extensive list of desiderata each
addressing different needs. To name but a few, we are interested in
generating counterfactuals that close
\protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable
\protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic
\protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible
causally founded
\protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

Efforts so far have largely been directed at improving the quality of
Counterfactual Explanations within a static context: given some
pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are
interested in generating one or multiple meaningful Counterfactual
Explanations for some individual characterized by \(x\). The ability of
Counterfactual Explanations to handle dynamics like data and model
shifts remains a largely unexplored research challenge at this point
\protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. We have been
able to identify only one recent work that considers the implications of
\textbf{exogenous} domain and model shifts in the context of AR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts
are strictly of external origin. For example, they might stem from data
correction, temporal shifts or geospatial changes
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a
framework for Algorithmic Recourse that evidently improves robustness to
such exogenous shifts.

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors
propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts --
situations where the distribution of data changes over time. Rabanser et
al.~suggest a framework to detect data drift from a minimal number of
samples through the application of two-sample tests
\protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a
generalization of the anomaly detection problem for large datasets,
which aims to answer the question if two sets of samples could have been
generated from the same probability distribution. Numerous approaches to
anomaly detection have been summarized
\protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another
well-established research topic is that of concept drift -- situations
where external variables influence the patterns between the input and
the output of a model
\protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama
et al.~offer a review of the adaptive learning techniques which can
handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}.
Less previous work is available on the related topic of model drift -
changes in model performance over time. Nelson et al.~review how
resistant different machine learning models are to the model drift
\protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et
al.~offer a method to detect changes in model performance when ground
truth is not available
\protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of Algorithmic Recourse, domain and model shifts were
first brought up by the authors behind ROAR
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work
they refer to model shifts as simply any perturbation \(\Delta\) to the
parameters of the model in question: \(M\). While this also sets the
baseline for our analysis here, it is worth noting that in
\protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these
perturbations are mechanically introduced. In contrast we are interested
in quantifying model shifts that arise endogenously as part of a dynamic
recourse process. In addition to quantifying the magnitude of shifts
\(\Delta\), we aim to also analyse the characteristics of changes to the
model, such as the position of the decision boundary and the overall
decisiveness of the model. We have not been able to identify previous
work on this topic.

@@

Despite the large and growing number of approaches to counterfactual
search, there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python
framework - CARLA - that can be used to apply and benchmark different
methodologies; finally, \texttt{CounterfactualExplanations.jl}
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits. In particular, we have built a
framework on top of \texttt{CounterfactualExplanations.jl} that extends
the functionality from static benchmarks to simulation experiments:
\texttt{AlgorithmicRecourseDynamics.jl}\footnote{The package is
  available from \ldots{}}. The core concepts implemented in that
package reflect what is presented in section Section~\ref{sec-method-2}
of this paper.

@@

We have chosen to focus on gradient-based counterfactual search for two
reasons: firstly, they can be seen as direct descendants of our baseline
method - Wachter; secondly, gradient-based search is particularly
well-suited for differentiable black-box models like deep neural
networks, which we focus on in this work. In particular, we include
include the following generators in our simulation experiments below:
\textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}},
\textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}},
\textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
and a greedy approach that relies on probabilistic models
\protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for
including these different generators in our analysis, is that they all
offer slightly different approaches to generate meaningful
counterfactuals for differentiable black-box models. We hypothesize that
generating more \textbf{meaningful} counterfactuals should mitigate the
endogenous dynamics illustrated in Figure~\ref{fig-poc} in
Section~\ref{sec-intro}. This intuition stems from the underlying idea
that more meaningful counterfactuals are generated by the same or at
least a very similar data generating process as the training data. All
else equal, counterfactuals that fulfill this basic requirement should
be less prone to trigger domain and model shifts.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked
\(K\)-dimensional array of counterfactual states and
\(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual
state space to the feature space. In Wachter, the state space is the
feature space: \(f\) is just the identity function and the number of
counterfactuals \(K\) is equal to one. Both REVISE and CLUE search
counterfactuals in some latent embedding \(S \subset \mathcal{S}\)
instead of the feature space directly. The latent embedding is learned
by a separate generative model that is tasked with learning the data
generating process (DGP) of \(X\). In this case \(f\) in
Equation~\ref{eq-general} corresponds to the decoder part of the
generative model, in other words the function that maps back from the
latent embedding to the feature space. Provided the generative model is
well-specified, traversing the latent embedding typically results in
realistic and plausible counterfactuals, because they are implicitly
generated by the (learned) DGP
\protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.

CLUE distinguishes itself from REVISE and other counterfactual
generators in that it aims to minimize the predictive uncertainty of the
model in question, \(M\). To quantify predictive uncertainty the authors
rely on entropy estimates for probabilistic models. The approach
proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which
we shall refer to as \textbf{Greedy}, also works with the subclass of
models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce
predictive uncertainty estimates. The authors show that in this setting
the cost function \(h(\cdot)\) in Equation~\ref{eq-general} is redundant
and meaningful counterfactuals can be generated in a fast and efficient
manner through a modified Jacobian-based Saliency Map Attack (JSMA).
Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also
show that by maximizing the predicted probability of \(x^\prime\) being
assigned to target class \(y^*\) we also implicitly minimize predictive
entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to
REVISE in the Bayesian context and we shall therefore refer to both
approaches collectively as \textbf{Latent Space} generators\footnote{In
  fact, there are a number of other recently proposed approaches to
  counterfactual search that also broadly fall in this same category.
  They largely differ with respect to the chosen generative model: for
  example, the generator proposed by
  \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies
  on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}}
distinguishes itself from all other generators considered here in that
it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et
al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that
diverse outcomes can in principal be achieved simply rerunning
counterfactual search multiple times using stochastic gradient descent
(or by randomly initializing the counterfactual). In
\protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is
explicitly proxied via Determinantal Point Processes (DDP): the authors
simply introduce DDP as a component of the cost function
\(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals
\(s_1, ... , s_K\) that look as different from each other as possible.
The implementation of DiCE in the our library of choice -
\texttt{CounterfactualExplanations.jl} - uses that exact approach. It is
worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is
constant and therefore does not affect the objective function
Equation~\ref{eq-general}.

@@

Below we first present our main findings with respect to the presence of
endogenous macrodynamics in Algorithmic Recourse. We then go on to
introduce and test a number of potential mitigation strategies.

@@

In light of the huge number of experiments we have conducted, we shall
start this section off with a few high-level observations. Across all
datasets (synthetic and real), classifiers and counterfactual generators
we observe either most or all of the following dynamics at varying
degrees:
